A White Noise Approach to Evolutionary Ecology

### 2020-04-06 .A White Noise Approach To Evolutionary Ecology/draft.Rmd
fixed a few typos

where $r\in\mathbb{R}$ is the intrinsic growth rate in the absence of abiotic selection, $a>0$ is the strength of stabilizing selection around the phenotypic optimum $\theta\in\mathbb{R}$, $c>0$ captures the sensitivity of fitness to intraspecific competition and we have set $\kappa(x,y)=1$. This exemplary fitness function has a few convenient properties that we breifly take note of. First, the effect of competition induces a local carrying capacity on the population, leading to a finite equilibrium abundance over bounded subsets of $\mathbb{R}$. Second, abiotic selection prevents the abundance density from diffusing too far from the abiotic optimum, leading to a finite equilibrial abundance across all of $\mathbb{R}$, $\lim_{t\to\infty}N(t)=\tfrac{1}{c}(r-\tfrac{1}{2}\sqrt{a\mu})$, finite equilibrial mean trait $\lim_{t\to\infty}\bar x(t)=\theta$ and finite equilibrial phenotypic variance $\lim_{t\to\infty}\sigma^2(t)=\sqrt{\mu/a}$, so long as $|\bar x(0)|<+\infty$, $\sigma^2(0),N(0)>0$ and $r\geq\sqrt{a\mu}/2$. We prove these statements in SM \S\ref{equilib}. By replacing $\kappa(x,y)=1$ with a particular Gaussian function, we will find in \S\ref{coev} and SM \S\ref{diffuse} this fitness function becomes equivalent to a fitness function derived from niche theory that we use to model coevolution driven by resource competition.

@@

Recall that $m(\nu,x)$ is shorthand for $m(\nu(x,t),x)$. That is, $m:[0,+\infty)\times\mathbb{R}\to\mathbb{R}$. Following our assumptions of the main text, we have that $m(y,x)$ is differentiable with respect to both $x$ and $y$ and there exists $r\in\mathbb{R}$ such that $m(y,x)\leq r$ across all $y\geq0$ and $x\in\mathbb{R}$. As in the main text, we also assume the initial condition $\nu(x,0)$ satisfies
\begin{equation}
N(0)=\int_{\mathbb{R}}\nu(x,0)dx<+\infty,
\end{equation}
\begin{equation}
-\infty<\bar x(0)=\int_{\mathbb{R}}xp(x,0)dx<+\infty,
\end{equation}
\begin{equation}
\sigma^2(0)=\int_{\mathbb{R}}(x-\bar x(0))^2p(x,0)dx<+\infty,
\end{equation}
where $p(x,0)=\nu(x,0)/N(0)$, and we consider the PDE
\begin{equation}\label{PDE_SM}
\dot\nu(x,t)=m(\nu,x)\nu(x,t)+\frac{\mu}{2}\Delta\nu(x,t).
\end{equation}

Replacing $m$ with it's upper bound $r\in\mathbb{R}$, equation (\ref{PDE_SM}) reduces to a simple parabolic equation that can be solved using standard techniques [@stanleyfarlow1993]. In particular, when $m(\nu,x)\equiv0$ denote the solution to (\ref{PDE_SM}) by $\nu_0(x,t)$. Then, denoting
\begin{equation}
\Phi(x,t)=\frac{\exp\left(-x^2/2\mu t\right)}{\sqrt{2\pi\mu t}},
\end{equation}
we have
\begin{equation}
\nu_0(x,t)=\int_{\mathbb{R}}\Phi(x-y,t)\nu(y,0)dy.
\end{equation}
In the more general case, when $m(\nu,x)\equiv r\in\mathbb{R}$, equation (\ref{PDE_SM}) has the solution $\nu_r(x,t)=e^{rt}\nu_0(x,t)$. Hence, $\nu_r(x,t)\geq0$ for all $x\in\mathbb{R}$ and $\int_\mathbb{R}\nu_r(x,t)dx=e^{rt}N(0)<+\infty$ for all $t\geq0$. Furthermore, denoting 
\begin{equation}
\begin{matrix}
N_r(t)&=&\int_\mathbb{R}\nu_r(x,t)dx, \\
p_r(x,t)&=&\nu_r(x,t)/N_r(t), \\
\bar x_r(t)&=&\int_\mathbb{R}xp_r(x,t)dx, \\
\sigma^2_r(t)&=&\int_\mathbb{R}(x-\bar x_r(t))^2p_r(x,t)dx, 
\end{matrix}
\end{equation}
we have
\begin{equation}
\bar x_r(t)=\int_\mathbb{R}x\int_\mathbb{R}\Phi(x-y,t)p(y,0)dydx=\int_\mathbb{R}yp(y,0)dy=\bar x(0),
\end{equation}
\begin{equation}
\sigma_r^2(t)=\int_\mathbb{R}(x-\bar x_r(t))^2\int_\mathbb{R}\Phi(x-y,t)p(y,0)dydx=\int_\mathbb{R}\Big((y-\bar x(0))^2+\mu t\Big)p(y,0)dy=\sigma^2(0)+\mu t.
\end{equation}
Hence, $|\bar x_r(t)|,\sigma_r^2(t)<+\infty$ for all $t\geq0$. For the sake of contradiction, suppose there exists $x\in\mathbb{R}$ and $t\geq0$ such that $\nu(x,t)>\nu_r(x,t)$. Then
\begin{equation}
\nu(x,t)-\nu(x,0)=\int_0^tm(\nu,x)\nu(x,s)+\frac{\mu}{2}\Delta\nu(x,s)ds>\int_0^tr\nu_r(x,s)+\frac{\mu}{2}\Delta\nu_r(x,s)ds=\nu_r(x,t)-\nu(x,0)
\end{equation}
which implies there exists $y\geq0$ and $x\in\mathbb{R}$ such that $m(y,x)>r$. But this contradicts our assumption $m(y,x)\leq r$ for all $y\geq0$ and $x\in\mathbb{R}$. So we have $\nu(x,t)\leq\nu_r(x,t)$ for each $x\in\mathbb{R}$ and $t\geq0$. This implies that $N(t)=\int_\mathbb{R}\nu(x,t)dx<+\infty$,
\begin{equation}
0\leq\int_\mathbb{R}x^2\nu(x,t)dx\leq\int_\mathbb{R}x^2\nu_r(x,t)dx<+\infty
\end{equation}
and in particular
\begin{equation}
0\leq\sigma^2(t)+\bar x^2(t)=\frac{1}{N(t)}\int_\mathbb{R}x^2\nu(x,t)dx<+\infty
\end{equation}
for each $t\geq0$.

@@

Here we set out to show, given appropriate initial conditions, $N_\infty=\lim_{t\to\infty}N(t)=\tfrac{1}{c}(r-\tfrac{1}{2}\sqrt{a\mu})$, $\bar x_\infty=\lim_{t\to\infty}\bar x(t)=\theta$ and $\sigma^2_\infty=\lim_{t\to\infty}\sigma^2(t)=\sqrt{\tfrac{\mu}{a}}$ when
\begin{equation}\label{m_log_stab}
m(\nu,x)=r-\frac{a}{2}(\theta-x)^2-c\int_\mathbb{R}\nu(y,t)dy
\end{equation}
for $\theta\in\mathbb{R}$, $\mu\geq0$, $a,c>0$ and $r\geq\tfrac{1}{2}\sqrt{\mu a}$. Set $\nu_\infty(x)=\lim_{t\to\infty}\nu(x,t)$ and $p_\infty(x)=\nu_\infty(x)/N_\infty$, when these quantities exist. The mean fitness can be calculated as
\begin{equation}
\bar m(t)=r-\frac{a}{2}((\theta-\bar x(t))^2-\sigma^2(t))-cN(t), \ \ \forall \ \ t\geq0.
\end{equation}
Solving for $N_\infty$ amounts to setting $\dot N(t)=0$ and solving for $N(t)$. This reduces to solving $\bar m(t)=0$ for $N(t)$, which returns
\begin{equation}
N_\infty=\frac{1}{c}\left(r-\frac{a}{2}\left((\theta-\bar x_\infty)^2-\sigma^2_\infty\right)\right).
\end{equation}
To find $\nu_\infty(x)$ we solve $\tfrac{\partial}{\partial t}\nu(x,t)=0$ for $\nu(x,t)$. This implies the following ordinary differential equation
\begin{equation}
\frac{d^2}{dx^2}\nu_\infty(x)=\left(\frac{2c}{\mu}N_\infty+\frac{a}{\mu}(\theta-x)^2-\frac{2r}{\mu}\right)\nu_\infty(x)
\end{equation}
which has the solution
\begin{equation}
\nu_\infty(x)=\frac{N_\infty}{\sqrt{2\pi}}\left(\frac{a}{\mu}\right)^{\frac{1}{4}}\exp\left(-\frac{(\theta-x)^2}{2\sqrt{\mu/a}}\right).
\end{equation}
From this expression we infer $\bar x_\infty=\theta$ and $\sigma^2_\infty=\sqrt{\mu/a}$. Hence $N_\infty=\frac{1}{c}\left(r-\frac{1}{2}\sqrt{a\mu}\right)$. This holds for any initial $\nu(\cdot,0)\in L^1(\mathbb{R})$ such that $\bar x(0)\in\mathbb{R}, \ \sigma^2(0),N(0)>0$.

>>>>

where $r\in\mathbb{R}$ is the intrinsic growth rate in the absence of abiotic selection, $a>0$ is the strength of stabilizing selection around the phenotypic optimum $\theta\in\mathbb{R}$, $c>0$ captures the sensitivity of fitness to intraspecific competition and we have set $\kappa(x,y)=1$. This exemplary fitness function has a few convenient properties that we breifly take note of. First, the effect of competition induces a local carrying capacity on the population, leading to a finite equilibrium abundance over bounded subsets of $\mathbb{R}$. Second, abiotic selection prevents the abundance density from diffusing too far from the abiotic optimum, leading to a finite equilibrial abundance across all of $\mathbb{R}$, $\lim_{t\to\infty}N(t)=\tfrac{1}{c}(r-\tfrac{1}{2}\sqrt{a\mu})$, finite equilibrial mean trait $\lim_{t\to\infty}\bar x(t)=\theta$ and finite equilibrial phenotypic variance $\lim_{t\to\infty}\sigma^2(t)=\sqrt{\mu/a}$, so long as $\bar x(0)\in\mathbb{R}$, $\sigma^2(0),N(0)\in(0,+\infty)$ and $r>\sqrt{a\mu}/2$. We prove these statements in SM \S\ref{equilib}. By replacing $\kappa(x,y)=1$ with a particular Gaussian function, we will find in \S\ref{coev} and SM \S\ref{diffuse} this fitness function becomes equivalent to a fitness function derived from niche theory that we use to model coevolution driven by resource competition.

@@

Recall that $m(\nu,x)$ is shorthand for $m(K(\nu)(x,t),x)$. That is, $m:[0,+\infty)\times\mathbb{R}\to\mathbb{R}$. Following our assumptions of the main text, we have that $m(h,x)$ is differentiable with respect to both $x$ and $h$ and there exists $r\in\mathbb{R}$ such that $m(h,x)\leq r$ across all $h\geq0$ and $x\in\mathbb{R}$. As in the main text, we also assume the initial condition $\nu(x,0)$ satisfies
\begin{equation}
N(0)=\int_{\mathbb{R}}\nu(x,0)dx<+\infty,
\end{equation}
\begin{equation}
-\infty<\bar x(0)=\int_{\mathbb{R}}xp(x,0)dx<+\infty,
\end{equation}
\begin{equation}
\sigma^2(0)=\int_{\mathbb{R}}(x-\bar x(0))^2p(x,0)dx<+\infty,
\end{equation}
where $p(x,0)=\nu(x,0)/N(0)$, and we consider the PDE
\begin{equation}\label{PDE_SM}
\dot\nu(x,t)=m(\nu,x)\nu(x,t)+\frac{\mu}{2}\Delta\nu(x,t).
\end{equation}

Replacing $m$ with it's upper bound $r\in\mathbb{R}$, equation (\ref{PDE_SM}) reduces to a simple parabolic equation that can be solved using standard techniques [@stanleyfarlow1993]. In particular, when $m(\nu,x)\equiv0$ denote the solution to (\ref{PDE_SM}) by $\nu_0(x,t)$. Then, denoting
\begin{equation}
\Phi(x,t)=\frac{\exp\left(-x^2/2\mu t\right)}{\sqrt{2\pi\mu t}},
\end{equation}
we have
\begin{equation}
\nu_0(x,t)=\int_{\mathbb{R}}\Phi(x-y,t)\nu(y,0)dy.
\end{equation}
In the more general case, when $m(\nu,x)\equiv r\in\mathbb{R}$, equation (\ref{PDE_SM}) has the solution $\nu_r(x,t)=e^{rt}\nu_0(x,t)$. Hence, $\nu_r(x,t)\geq0$ for all $x\in\mathbb{R}$ and $\int_\mathbb{R}\nu_r(x,t)dx=e^{rt}N(0)<+\infty$ for all $t\geq0$. Furthermore, denoting 
\begin{equation}
\begin{matrix}
N_r(t)&=&\int_\mathbb{R}\nu_r(x,t)dx, \\
p_r(x,t)&=&\nu_r(x,t)/N_r(t), \\
\bar x_r(t)&=&\int_\mathbb{R}xp_r(x,t)dx, \\
\sigma^2_r(t)&=&\int_\mathbb{R}(x-\bar x_r(t))^2p_r(x,t)dx, 
\end{matrix}
\end{equation}
we have
\begin{equation}
\bar x_r(t)=\int_\mathbb{R}x\int_\mathbb{R}\Phi(x-y,t)p(y,0)dydx=\int_\mathbb{R}yp(y,0)dy=\bar x(0),
\end{equation}
\begin{equation}
\sigma_r^2(t)=\int_\mathbb{R}(x-\bar x_r(t))^2\int_\mathbb{R}\Phi(x-y,t)p(y,0)dydx=\int_\mathbb{R}\Big((y-\bar x(0))^2+\mu t\Big)p(y,0)dy=\sigma^2(0)+\mu t.
\end{equation}
Hence, $|\bar x_r(t)|,\sigma_r^2(t)<+\infty$ for all $t\geq0$. For the sake of contradiction, suppose there exists $x\in\mathbb{R}$ and $t\geq0$ such that $\nu(x,t)>\nu_r(x,t)$. Then
\begin{equation}
\nu(x,t)-\nu(x,0)=\int_0^tm(\nu,x)\nu(x,s)+\frac{\mu}{2}\Delta\nu(x,s)ds>\int_0^tr\nu_r(x,s)+\frac{\mu}{2}\Delta\nu_r(x,s)ds=\nu_r(x,t)-\nu(x,0)
\end{equation}
which implies there exists $h\geq0$ and $x\in\mathbb{R}$ such that $m(h,x)>r$. But this contradicts our assumption $m(h,x)\leq r$ for all $h\geq0$ and $x\in\mathbb{R}$. So we have $\nu(x,t)\leq\nu_r(x,t)$ for each $x\in\mathbb{R}$ and $t\geq0$. This implies that $N(t)=\int_\mathbb{R}\nu(x,t)dx<+\infty$,
\begin{equation}
0\leq\int_\mathbb{R}x^2\nu(x,t)dx\leq\int_\mathbb{R}x^2\nu_r(x,t)dx<+\infty
\end{equation}
and in particular
\begin{equation}
0\leq\sigma^2(t)+\bar x^2(t)=\frac{1}{N(t)}\int_\mathbb{R}x^2\nu(x,t)dx<+\infty
\end{equation}
for each $t\geq0$.

@@

Here we set out to show, given appropriate initial conditions, $N_\infty=\lim_{t\to\infty}N(t)=\tfrac{1}{c}(r-\tfrac{1}{2}\sqrt{a\mu})$, $\bar x_\infty=\lim_{t\to\infty}\bar x(t)=\theta$ and $\sigma^2_\infty=\lim_{t\to\infty}\sigma^2(t)=\sqrt{\tfrac{\mu}{a}}$ when
\begin{equation}\label{m_log_stab}
m(\nu,x)=r-\frac{a}{2}(\theta-x)^2-c\int_\mathbb{R}\nu(y,t)dy
\end{equation}
for $\theta\in\mathbb{R}$, $\mu\geq0$, $a,c>0$ and $r>\tfrac{1}{2}\sqrt{\mu a}$. When it exists, set $\nu_\infty(x)=\lim_{t\to\infty}\nu(x,t)$. The mean fitness can be calculated as
\begin{equation}
\bar m(t)=r-\frac{a}{2}((\theta-\bar x(t))^2-\sigma^2(t))-cN(t), \ \ \forall \ \ t\geq0.
\end{equation}
Solving for $N_\infty$ amounts to setting $\dot N(t)=0$ and solving for $N(t)$. This reduces to solving $\bar m(t)=0$ for $N(t)$, which returns
\begin{equation}
N_\infty=\frac{1}{c}\left(r-\frac{a}{2}\left((\theta-\bar x_\infty)^2-\sigma^2_\infty\right)\right).
\end{equation}
To find $\nu_\infty(x)$ we solve $\tfrac{\partial}{\partial t}\nu(x,t)=0$ for $\nu(x,t)$. This implies the following ordinary differential equation
\begin{equation}
\frac{d^2}{dx^2}\nu_\infty(x)=\left(\frac{2c}{\mu}N_\infty+\frac{a}{\mu}(\theta-x)^2-\frac{2r}{\mu}\right)\nu_\infty(x)
\end{equation}
which has the solution
\begin{equation}
\nu_\infty(x)=\frac{N_\infty}{\sqrt{2\pi}}\left(\frac{a}{\mu}\right)^{\frac{1}{4}}\exp\left(-\frac{(\theta-x)^2}{2\sqrt{\mu/a}}\right).
\end{equation}
From this expression we infer $\bar x_\infty=\theta$ and $\sigma^2_\infty=\sqrt{\mu/a}$. Hence $N_\infty=\frac{1}{c}\left(r-\frac{1}{2}\sqrt{a\mu}\right)$. This holds for any initial $\nu(\cdot,0)\in L^1(\mathbb{R})$ such that $\bar x(0)\in\mathbb{R}, \ \sigma^2(0),N(0)\in(0,+\infty)$.
